On Supporting Molecular Visualization on Grid Resources using Visual Programming Environments

### 2008-04-24 opal/documents/grid2008/paper.tex
added a bunch of references

  The visual programming paradigm provided by tools such as AVS [ref],
  OpenDX [ref], SciRun [ref], etc have been around for a couple decades.
  Recently, with the advent of the Web 2.0 paradigm and the concept of {\em
    mashups} [ref], there has been renewed interest in such types of
  environments. Over the past several years, we have developed one such
  tool called Vision [ref], which is a component-based, application-domain
  agnostic, cross-platform visual programming environment. Meanwhile,
  advances in distributed and Grid computing have opened up avenues for
  scientists to run large simulations on a scale that was impossible until
  recent times. In the past, visual programming tools were generally
  limited to running scientific experiments on user's individual
  workstations, and had limited capability to use distributed resources.
  With the proliferation of Web services, distributed resources are much
  more easy to access from any workstation that is connected to the
  Internet. In particular, we have developed the Opal toolkit, which
  provides a mechanism to wrap scientific applications running on Grid
  resources as Web services, thus providing the clients with a simple Web
  API to the back-end computational functionality. Opal encapsulates
  standard Grid security mechanisms, schedulers, and state management for
  jobs, and also provides a mechanism to describe command-line arguments
  that can be used to automatically generate user-interfaces [ref]. In this
  paper, we describe our current work on enabling access to scientific
  applications running on Grid resources via Opal from the Vision
  environment. We present brief overviews of both Vision and Opal, and
  discuss a scientific workflow that has been implemented with the above
  tools. We discuss our experiences, and plans for future work.

@@

The visual-programming paradigm provided by tools such as the AVS [X],
OpenDX [X], Khoros [X], DataExplorer [X], Ptoleme [X] or Scirun, [X] has
been around for a couple of decades. Recently, there has recently been a
renewed interest in this type of user interface and numerous pipelining
tools have become available, including: PHENIX [x], pipeline pilot [X],
mapero [X], Orange [X], Elefant [X], Kepler [X], Taverna [X] to name a few.

@@

Over the past 8 years we have developed Vision [X], a component-based,
application-domain agnostic, cross-platform visual programming environment.
Vision has been designed as an extension to the Python programming language
[X] thus providing a fully fledged, high level, object-oriented,
interpretive language for the interactive manipulation of data and/or
modification of the application itself.

@@

To date, several libraries of nodes are available [Fig X]. The standard
library comprises nodes exposing Python keywords (print, eval, setattr and
call methods), along with nodes enabling the generation of simple input
data such as floats, integers, and strings. The 3-D Visualization library
is based on the DejaVu component and provides nodes to create 3-D
geometries and display them. The SymServ library implements a set of nodes
defining geometric transformations such as symmetries, translation,
rotation, helical arrangements etc. that can be combined to created streams
of nested 3-D transformations. The MolKit library is based on the MolKit
package and provides nodes to read molecular data files and manipulate
molecules. The Imaging library currently exposes a subset of the
functionality available in the Python Image Library (PIL) thus enabling the
creation of image processing pipelines.

@@

Several tools have been developed by the scientific community to support
the design and execution of workflows. Each one of them use different
approaches to model a graphical environment for designing and executing
workflow.  In this section we present three workflow management systems:
Kepler, Taverna, and AVS. This paragraph does not want to provide an
exaustive survey of these systems, for more information refer to [grid].

Kepler [6] is a workflow system developed under several research projects.
It is written in Java and based on Ptolemy II [6], a system originally
created to study modeling, simulation, and design of concurrent systems.
Kepler models workflow in MoMl, a markup language developed in Ptolemy that
is based on the concept of actors. Each actor has input and output ports
and wraps typical workflow actions like invoke a Web Service or a Grid
Services, launch a Globus Grid Jobs, perform a GridFTP transfer, and many
others. The designer can model non Directed Acyclic Graph [5] using a
simple Graphic User Interface where the user has simply to drag and drop
icons and connect them.  Kepler can also be classified as a workflow system
based on the data flow model [6], where data is streamed from one actor to
another triggering their activation. Moreover an extensible architecture
based on the plugin model allows developers to easily interagte thrid party
tools.

Taverna is an open source software tool for designing and executing
bioinformatics workflows, developed in the myGrid project [7]. It uses
Scufl as a runtime language for the execution of the workflow, and Freefluo
as an execution engine. Scufl can be extended using Processor plugins that
manage the interaction with different external service interfaces. Taverna
is shipped with a set of processors that guarantee interaction with
services based on WS-I standard. The designer is based on a directed ciclic
graph that represents data flow, which uses a simple graphical notation
developed by the Taverna team. The GUI can be easily extended creating new
palettes and new widgets with development of new plugins.

@@

Programming and using distributed Grid resources is quite a challenge task.
There is hope that Service-Oriented Architectures (SOA) can address this
problem by providing platform-independent, language-neutral service
interfaces that hide the complexity of the implementations while providing
a well-defined and high-performance Quality of Service (QoS). In our recent
paper [ref], we claimed that in order to lower the entrance barrier for
scientific application developers and end-users to Grid infrastructures, we
must do the following:

% TODO: possibly expand every bullet (04/23)
\begin{enumerate}
\item{Focus on application-level services}
\item{Provide users access to virtualized resources}
\item{Provide multiple user interfaces to scientific users}
\item{Lower deployment costs}
\item{Follow the Web 2.0 paradigm by enabling user participation and contribution}
\end{enumerate}

Towards this end, we have developed the Opal toolkit
\cite{opal} for wrapping scientific applications as Web services. In order
to wrap an application as a Web service, application developers are
expected to write a simple XML-based application configuration. The
application configuration contains information about the scientific
application, such as application name, binary location, and application
metadata (e.g. usage information). Using a simple Apache Ant task, the Opal
toolkit can deploy the application as a Web service into a container based
on Apache Tomcat and Axis. Once the application is deployed as a Web
service, clients can access this service programmatically using its WSDL
description. The WSDL API provides operations for job launch (which accepts
command-line arguments and input files as its parameters), querying status,
and retrieving outputs. Furthermore, it also provides an API for retrieving
application metadata. WSDL savvy users could write their own clients to
access Opal based applications. However, not every scientific user is
capable or interested in writing Web service clients. Hence, several
interfaces have been provided for the end-user. Apart from command-line
clients written in languages like Java, Python [4] and Perl [15] available
from NBCR, Web interfaces for Opal based services are available through the
GridSphere portal environment [18], such as the MEME portlet in My
WorkSphere [24]. In addition, Opal based interfaces via Rich Internet
Applications (RIA) such as Gemstone [20], and Problem Solving Environments
(PSE) such as the Python Molecular Viewer (PMV) [22], AutoDockTools [21]
and Continuity [27] have also been provided.

@@

\begin{figure}
\begin{center}
\includegraphics[width=8.5cm]{opal2-providers}
\end{center}
\caption{The Opal2 Provider Model}
\label{fig:opal2-providers}
\end{figure}

@@

%% References
% TODO (04/23)
\bibliographystyle{plain}
\bibliography{paper}

>>>>

  The visual programming paradigm provided by tools such as AVS \cite{avs},
  OpenDX \cite{opendx}, SciRun \cite{scirun}, etc have been around for a
  couple decades. Recently, with the advent of the Web 2.0 paradigm and the
  concept of {\em mashups} \cite{web2.0}, there has been renewed interest
  in such types of environments. Over the past several years, we have
  developed one such tool called Vision \cite{viper}, which is a
  component-based, application-domain agnostic, cross-platform visual
  programming environment. Meanwhile, advances in distributed and Grid
  computing have opened up avenues for scientists to run large simulations
  on a scale that was impossible until recent times. In the past, visual
  programming tools were generally limited to running scientific
  experiments on user's individual workstations, and had limited capability
  to use distributed resources. With the proliferation of Web services,
  distributed resources are much more easy to access from any workstation
  that is connected to the Internet. In particular, we have developed the
  Opal toolkit, which provides a mechanism to wrap scientific applications
  running on Grid resources as Web services, thus providing the clients
  with a simple Web API to the back-end computational functionality. Opal
  encapsulates standard Grid security mechanisms, schedulers, and state
  management for jobs, and also provides a mechanism to describe
  command-line arguments that can be used to automatically generate
  user-interfaces \cite{opal}. In this paper, we describe our current work
  on enabling access to scientific applications running on Grid resources
  via Opal from the Vision environment. We present brief overviews of both
  Vision and Opal, and discuss a scientific workflow that has been
  implemented with the above tools. We discuss our experiences, and plans
  for future work.

@@

The visual-programming paradigm provided by tools such as the AVS
\cite{avs}, OpenDX \cite{opendx}, or SciRun \cite{scirun} has been around
for a couple of decades. Recently, there has recently been a renewed
interest in this type of user interface and numerous pipelining tools have
become available, including: PHENIX [ref], pipeline pilot [ref], mapero
[ref], Orange [ref], Elefant [ref], Kepler \cite{kepler}, Taverna
\cite{taverna} to name a few.

@@

Over the past 8 years we have developed Vision \cite{viper}, a
component-based, application-domain agnostic, cross-platform visual
programming environment. Vision has been designed as an extension to the
Python programming language thus providing a fully fledged, high level,
object-oriented, interpretive language for the interactive manipulation of
data and/or modification of the application itself.

@@

To date, several libraries of nodes are available as seen in Figure
\ref{fig:vision-overview}. The standard library comprises nodes exposing
Python keywords (print, eval, setattr and call methods), along with nodes
enabling the generation of simple input data such as floats, integers, and
strings. The 3-D Visualization library is based on the DejaVu component and
provides nodes to create 3-D geometries and display them. The SymServ
library implements a set of nodes defining geometric transformations such
as symmetries, translation, rotation, helical arrangements etc. that can be
combined to created streams of nested 3-D transformations. The MolKit
library is based on the MolKit package and provides nodes to read molecular
data files and manipulate molecules. The Imaging library currently exposes
a subset of the functionality available in the Python Image Library (PIL)
thus enabling the creation of image processing pipelines.

@@

Several tools have been developed by the scientific community to support
the design and execution of workflows. Each one of them use different
approaches to model a graphical environment for designing and executing
workflow. In this section we present three workflow management systems:
Kepler, Taverna, and AVS. This paragraph does not want to provide an
exaustive survey of these systems, for more information refer to
\cite{wf-grid}.

Kepler is a workflow system developed under several research projects. It
is written in Java and based on Ptolemy II \cite{ptolemy}, a system
originally created to study modeling, simulation, and design of concurrent
systems. Kepler models workflow in MoMl, a markup language developed in
Ptolemy that is based on the concept of actors. Each actor has input and
output ports and wraps typical workflow actions like invoke a Web Service
or a Grid Services, launch a Globus Grid Jobs, perform a GridFTP transfer,
and many others. The designer can model non Directed Acyclic Graph using a
simple Graphic User Interface where the user has simply to drag and drop
icons and connect them. Kepler can also be classified as a workflow system
based on the data flow model \cite{kepler}, where data is streamed from one
actor to another triggering their activation. Moreover an extensible
architecture based on the plugin model allows developers to easily
interagte thrid party tools.

Taverna is an open source software tool for designing and executing
bioinformatics workflows, developed in the myGrid project \cite{taverna}.
It uses Scufl as a runtime language for the execution of the workflow, and
Freefluo as an execution engine. Scufl can be extended using Processor
plugins that manage the interaction with different external service
interfaces. Taverna is shipped with a set of processors that guarantee
interaction with services based on WS-I standard. The designer is based on
a directed ciclic graph that represents data flow, which uses a simple
graphical notation developed by the Taverna team. The GUI can be easily
extended creating new palettes and new widgets with development of new
plugins.

@@

Programming and using distributed Grid resources is quite a challenge task.
There is hope that Service-Oriented Architectures (SOA) can address this
problem by providing platform-independent, language-neutral service
interfaces that hide the complexity of the implementations while providing
a well-defined and high-performance Quality of Service (QoS). In our recent
paper \cite{escience}, we claimed that in order to lower the entrance
barrier for scientific application developers and end-users to Grid
infrastructures, we must do the following:

\begin{enumerate}
\item{{\bf Focus on application-level services:} From the perspective of a
    domain scientist, the services that are most relevant are the ones that
    provide a scientific operation, and where the semantics of the
    operations are defined in terms of the domain science, and not computer
    science. This is what Foster terms ''services-oriented science''
    \cite{sos}.}

\item{{\bf Provide users access to virtualized resources:} The use of
    virtualized resources through a layer of indirection allows the system
    developers to add or move resources as needed without any interruption.}

\item{{\bf Provide multiple user interfaces to scientific users:} A single
    user interface is typically not sufficient for all the users using Grid
    resources. We advocate application services that can be used by a
    number of end-user tools, including Web-based portals, rich desktop
    applications, command-line tools and language-specific libraries.}

\item{{\bf Lower deployment costs:} Scientific end-users don’t have the
    time or money to invest in trying to figure out the complexity of the
    back-end resources, and deploying an application on Grid resources
    should be as easy as possible.}

\item{{\bf Follow the Web 2.0 paradigm by enabling user participation and
      contribution:} In principle, the Web 2.0 paradigm allows users to
    contribute their applications, share data, and participate in
    collaborative activities with ease and keep their focus on the process
    of creation, instead of maintaining and learning new tools and
    interfaces.}
\end{enumerate}

Towards this end, we have developed the Opal toolkit \cite{opal} for
wrapping scientific applications as Web services. In order to wrap an
application as a Web service, application developers are expected to write
a simple XML-based application configuration. The application configuration
contains information about the scientific application, such as application
name, binary location, and application metadata (e.g. usage information).
Using a simple Apache Ant task, the Opal toolkit can deploy the application
as a Web service into a container based on Apache Tomcat and Axis. Once the
application is deployed as a Web service, clients can access this service
programmatically using its WSDL description. The WSDL API provides
operations for job launch (which accepts command-line arguments and input
files as its parameters), querying status, and retrieving outputs.
Furthermore, it also provides an API for retrieving application metadata.
WSDL savvy users could write their own clients to access Opal based
applications. However, not every scientific user is capable or interested
in writing Web service clients. Hence, several interfaces have been
provided for the end-user. Apart from command-line clients written in
languages like Java, Python and Perl available from NBCR, Web interfaces
for Opal based services are available through the GridSphere portal
environment \cite{gridsphere}, such as the MEME portlet in My WorkSphere
\cite{myworksphere}. In addition, Opal based interfaces via Rich Internet
Applications (RIA) such as Gemstone \cite{gemstone}, and Problem Solving
Environments (PSE) such as the Python Molecular Viewer (PMV), AutoDockTools
\cite{pmv-vision-adt} and Continuity \cite{continuity} have also been provided.

@@

\begin{figure}
\begin{center}
\includegraphics[width=9cm]{opal2-providers}
\end{center}
\caption{The Opal2 Provider Model}
\label{fig:opal2-providers}
\end{figure}

@@

%% References
\bibliographystyle{plain}
\bibliography{paper}
